Temperature, wave exposure and wind as drivers of morphology in two kelp species

### 2019-07-07 paper_morph_new.Rmd
pdf version added, word version updated and general house cleaning.

---
title: Temperature, wave exposure and wind as drivers of morphology in two kelp species
  along the coast of South Africa
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
output:
  word_document: 
    fig_caption: yes
tables: yes
fontsize: 10pt
papersize: A4
geometry: margin=1in  
---

@@

Wave exposure and temperature have also been shown to be the main drivers of morphological adaptation across various kelp species. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility (Denny & Gaylord 2002). For example, a study by (Wernberg & Thomsen 2005)) examined the consistency of wave exposure as a driver of *Ecklonia radiata* (C. Agardh) J. Agardh across a broad geographic range, and  showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites. The morphological adaptation to wave exposure must be kept in balance with other important processes, such as nutrient assimilation and light absorption, which are dependent on the amount of surface area of the blades. Therefore, there is a trade-off between reducing overall drag and maintaining nutrient and photosynthetic ability. Temperature has also been shown to play a role in driving kelp morphology. For example, a study by (Serisawa et al. 2002) compared the morphology of Ecklonia cava Kjellman growing in warm temperate and cool temperate morphologies. The results showed that wrinkles in the blade seem to be a characteristic of warm temperate regions. However this study did not take into account the interaction with other environmental variables such as wave exposure, which also affect kelp morphology. The morphological adapatibility of kelps is driven by a combination of environmental factors that in turn do not act independently of one another. A study by (Wernberg et al. 2003) investigated the morphology of *E. radiata* in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australasian coast. They found no correlation between spatial distance and morphological similarity and rather the morphology of kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales (Wernberg et al. 2003). Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Wave exposure may be greater in shallower environments, or it may be reduced through damping effects of kelp further offshore and natural barriers such as rock outcrops; also, seawater temperature may be higher in shallow water environments due to higher solar irradiance. The kelp species

Other species of kelp, such as *Laminaria pallida* and *Ecklonia maxima* (Osbeck) Papenfuss which are important habitat forming seaweeds that exist around the coast of South Africa, offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between deep and shallow water environments. Although both these species exist in the subtidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in the shallower waters. Therefore, *E. maxima* is exposed to variations in wave, swell and temperature, while *L. pallida* is exposed to variations in waves and temperature. (Molloy & Bolton 1996) investigated the effect of depth and wave exposure on *L. pallida* at different depths and wave exposure and  showed that depth had a greater effect than wave exposure when considering all the morphological characteristics; when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by (Rothman et al. 2017) investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. They suggested that turbidity was the environmental driver responsible for this change. Research using advanced numerical tools is lacking. Previous research investigating drivers of kelp morphological characteristics used qualitative estimates of wave exposure and did not consider other morphological characteristics. Furthermore, only shallow populations have been considered in previous research which ignores the effects of wave damping. No work currently exists where other morphological characters have been considered, quantitative estimates of environmental drivers have been calculated, and the addition of wind as a possible driver of morphological characteristics has been considered.

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of wave exposure. Cartographical models of wave exposure are regarded as 'fetch-based-models' which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modelling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitave, reproducible approach for measuring wave exposure. Currently, no ecological studies investigating macroalgal communities have incorporated 3D spectral numerical modelling. The aim of this study is, therefore, to understand how temperature, wave exposure and wind can influence the morphology in two species of kelps around South Africa. This will be achieved by initially understanding the variation in temperature and waves and the consequences for morphological characteristics of *E. maxima* and *L. pallida*.

@@

```{r echo=FALSE, fig.cap='Schematic of *E. maxima* on the left and *L. pallida* on the right.REF?'}

@@

```{r Temperature variable processing}

@@

```{r Temperature variables plot, echo=FALSE, fig.cap='Temperature variables at the collection sites around the Cape Peninsula. Temperature variables include minimum represented by blue dots, fig.height=9, fig.width=7}

@@

```{r Wave data prep, echo=FALSE}

@@

```{r Wave variable processing, echo=FALSE}

@@

```{r Wave variable plot, echo=FALSE, fig.cap='Wave variables at each collection site around the Western Cape coast. Mean variables are represented by dots and standard deviation by whiskers. Plot "A" represents annual significant wave height, fig.height=8, fig.width=8}

>>>>

---
title: Temperature, wave exposure and wind as drivers of morphology in two kelp species
  along the coast of South Africa
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  pdf_document: default
  word_document:
    fig_caption: yes
papersize: A4
tables: yes
fontsize: 10pt
---

@@

Wave exposure and temperature have also been shown to be the main drivers of morphological adaptation across various kelp species. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility (Denny & Gaylord 2002). For example, a study by Wernberg & Thomsen (2005) examined the consistency of wave exposure as a driver of *Ecklonia radiata* (C. Agardh) J. Agardh across a broad geographic range, and  showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites. The morphological adaptation to wave exposure must be kept in balance with other important processes, such as nutrient assimilation and light absorption, which are dependent on the amount of surface area of the blades. Therefore, there is a trade-off between reducing overall drag and maintaining nutrient and photosynthetic ability. Temperature has also been shown to play a role in driving kelp morphology. For example, a study by Serisawa et al. (2002) compared the morphology of Ecklonia cava Kjellman growing in warm temperate and cool temperate morphologies. The results showed that wrinkles in the blade seem to be a characteristic of warm temperate regions. However this study did not take into account the interaction with other environmental variables such as wave exposure, which also affect kelp morphology. The morphological adapatibility of kelps is driven by a combination of environmental factors that in turn do not act independently of one another. A study by Wernberg et al. (2003) investigated the morphology of *E. radiata* in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australasian coast. They found no correlation between spatial distance and morphological similarity and rather the morphology of kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales (Wernberg et al. 2003). Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Wave exposure may be greater in shallower environments, or it may be reduced through damping effects of kelp further offshore and natural barriers such as rock outcrops; also, seawater temperature may be higher in shallow water environments due to higher solar irradiance.

Other species of kelp, such as *Laminaria pallida* and *Ecklonia maxima* (Osbeck) Papenfuss which are important habitat forming seaweeds that exist around the coast of South Africa, offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between deep and shallow water environments. Although both these species exist in the subtidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in the shallower waters. Therefore, *E. maxima* is exposed to variations in wave, swell and temperature, while *L. pallida* is exposed to variations in waves and temperature. (Molloy & Bolton 1996) investigated the effect of depth and wave exposure on *L. pallida* at different depths and wave exposure and  showed that depth had a greater effect than wave exposure when considering all the morphological characteristics; when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by Rothman et al. (2017) investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. They suggested that turbidity was the environmental driver responsible for this change.

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of wave exposure. Cartographical models of wave exposure are regarded as 'fetch-based-models' which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modelling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitave, reproducible approach for measuring wave exposure. Currently, no ecological studies investigating macroalgal communities have incorporated 3D spectral numerical modelling. Previous research investigating drivers of kelp morphological characteristics used qualitative estimates of wave exposure and did not consider other morphological characteristics. Furthermore, only shallow populations have been considered in previous research which ignores the effects of wave damping. No work currently exists where other morphological characters have been considered, quantitative estimates of environmental drivers have been calculated, and the addition of wind as a possible driver of morphological characteristics has been considered. The aim of this study is, therefore, to understand how temperature, wave exposure and wind can influence the morphology in two species of kelps around South Africa. This will be achieved by initially understanding the variation in temperature and waves and the consequences for morphological characteristics of *E. maxima* and *L. pallida*.

@@

```{r echo=FALSE, fig.cap='Schematic of *E. maxima* on the left and *L. pallida* on the right. (Dyer 2018)'}

@@

```{r Temperature variable processing, message=FALSE, warning=TRUE, include=FALSE}

@@

```{r Temperature variables plot, echo=FALSE, fig.cap="Temperature variables at the collection sites around the Cape Peninsula. Temperature variables include minimum represented by blue dots.", fig.height=9, fig.width=7}

@@

```{r Wave data prep, message=FALSE, warning=FALSE, include=FALSE}

@@

```{r Wave variable processing, include=FALSE}

@@

```{r Wave variable plot, echo=FALSE, fig.cap='Wave variables at each collection site around the Western Cape coast. Mean variables are represented by dots and standard deviation by whiskers. Plot "A" represents annual significant wave height.', fig.height=8, fig.width=8}
